---
title: "Gene Expression Exploratory Analysis"
author: 'Felix E. Rivera-Mariani, PhD'
date: "August 7, 2016"
output: pdf_document
---

### Summary
This report present an exploratory data analysis for the gene counts for the 6 samples evaluated this capstone project: 3 samples correspond to fetus and 3 to adults. Among the exploratory analysis performed are distances between samples, boxplots of the FPKMs (both untransformed and log2 transformed), as well as PCA analysis. 

```{r, working directory, echo=FALSE}
setwd("C:/Users/Felix/Dropbox/Coursera/Genomic-Data-Science/8-Genomic-datascience-capstone/github-files")
```

```{r, datasets, echo=FALSE, warning=FALSE, message=FALSE}
# load datasets
## gene expression dataset
gene_exp <- read.delim("gene_approx_counts_in_FPKM.txt")

## dataset with phenotypic information
phenotypic <- read.delim("gdc-sample-phenotypic-summalign.txt")

## create tidy dataset of gene expression
library (tidyr)
gene_exp_long <- gather (gene_exp, sample, fpkm, FPKM535:FPKM566)

## re-order the sample column

gene_exp_long$sample <- factor (gene_exp_long$sample, order=TRUE, levels=c("FPKM534", "FPKM535",
	"FPKM538", "FPKM541", "FPKM561", "FPKM566"))
```

```{r, distance, echo=FALSE, warning=FALSE, message=FALSE}
sampleDists <- dist( t( (gene_exp[2:7])))

sampleDists
```

From the distance matrix above, runs 534, 535, and 561 are have the shortest distances (less than 10^-2). All other distances are higher than 10^5.  

```{r, boxplot, echo=FALSE, warning=FALSE, message=FALSE}
# boxplot of counts FPKM without transformation
boxplot (fpkm ~ sample, gene_exp_long, main= "Boxplots of FPKMs per Sample",
         ylab="FPKM", xlab="Sample", las=1)
```

The boxplot of the FPKMs per sample provides evidence that the FPKMs values must be transformed. Log2 transformation was selected. The boxplot for the log2 transformed FPKM per sample is found below. 

```{r, log2 transf, echo=FALSE, warning=FALSE, message=FALSE}
library (dplyr)
gene_exp_long <- mutate (gene_exp_long, logstrnf = log2(gene_exp_long$fpkm))

## boxplot of log2 transformed FPKM
boxplot(logstrnf ~ sample, gene_exp_long, main="Boxplots of Log2 Transformed FPKM per Sample", ylab="Log2 FPKM", xlab="Samples", las=1)
```

From the log2 boxplot above, the distribution of the FPKMs corresponding to the adult samples (534, 535, 561) shift a little higher than the distribution of the fetus samples (538, 541, 566)

To further evaluate the distances between samples, a PCA scatterplot was constructed.

```{r, PCA, echo=FALSE, message=FALSE, warning=FALSE}
# Principal Commponent Analysis
## calculate the mean, including all fpkm
mean_all <- mean (gene_exp_long$fpkm)

## substract the mean of fpkms to each fpkm observation
gene_exp_long <- mutate (gene_exp_long, dfallmean = mean_all - gene_exp_long$fpkm)

## calculat the mean difference by run and create a data frame
pca <- gene_exp_long %>% group_by(sample) %>% summarise(dfmeanall_fpkm = mean(dfallmean))
pca <- as.data.frame(pca)

## calculate the mean fpkm for each sample
samples <- select(gene_exp, FPKM534, FPKM535, FPKM538, FPKM541, FPKM561, FPKM566)
mean_all_within <- apply(samples, 2, mean)

## create columns in which the mean fpkm for each sample is substracted to each sample
gene_exp <- mutate (gene_exp, FPKM534mindiff = FPKM534 - 28.5, FPKM535mindiff = FPKM535 - 28.5, FPKM538mindiff = FPKM538 - 56.4, FPKM541mindiff = FPKM541 - 56.4, FPKM561mindiff = FPKM561 - 14.4, FPKM566mindiff = FPKM566 - 37.5)

## calculate the mean for each sample mean difference columns 
gene_exp_mindiff_within <- select(gene_exp, FPKM534mindiff, FPKM535mindiff, FPKM538mindiff, FPKM541mindiff, FPKM561mindiff, FPKM566mindiff)
meandiff_within <- apply (gene_exp_mindiff_within, 2, mean)

## add the phenotypic variables
pca <- data.frame (pca, meandiff_within)
age <- c("adult", "adult", "fetus", "fetus", "adult", "fetus")
pca <- data.frame (pca, meandiff_within, age)
sample_id <- c("R2857", "R2869", "R3462", "R3485", "R4166", "R4706")
pca <- data.frame (pca, meandiff_within, age, sample_id)

## create a pca scatter plot
library (ggplot2)
pcaplot <- ggplot(pca, aes(meandiff_within, dfmeanall_fpkm, color=age, shape=sample)) + geom_point(size=3) + xlab("Variance of Log2 FPKM withn Samples") + ylab("Overall FPKM Log2 Variance")
pcaplot
```

The PCA scatter plot above supports the distance matrix mentionved earlier in this report. In the scatter plot above, FPKMs distances corresponding to fetus samples (538, 541, and 566) form a cluster, and those for adult samples (534, 535, 561) form another although sample 561 seems a bit closer to the fetus than the adult cluster.  
